---
title: "AIC, BIC, and Deviance"
author: "Luke Miratrix"
editor: 
  markdown: 
    wrap: sentence
---

```{r, include=FALSE}
library( blkvar )
```

In this section, we briefly walk through how to find AIC, BIC, and Deviance to compare models.
We have a simple multilevel dataset (we generate through a utility package, `blkvar`, that Miratrix and the C.A.R.E.S. lab has used to explore how multilevel modeling works in practice), and generate a few variables that we will use as predictors.
Only the fourth variable is actually useful for prediction!
Let's see if our AIC, etc., measures identify which model is superior.

To install a "working package" we use `devtools`:

```{r, eval=FALSE}
devtools::install_github("https://github.com/lmiratrix/blkvar" )
```

```{r}
library( blkvar )
dd = generate_multilevel_data( J = 40 )
head( dd )

dd$X1 = rnorm( nrow(dd) )
dd$X2 = rnorm( nrow(dd) )
dd$X3 = rnorm( nrow(dd) )
dd$X4 = dd$Yobs + rnorm(nrow(dd))
 
M1 = lmer( Yobs ~ 1 + (1|sid), data=dd )
M2 = lmer( Yobs ~ 1 + X1 + X2 + X3 + (1|sid), data=dd )
M3 = lmer( Yobs ~ 1 + X1 + X2 + X3 + X4 + (1|sid), data=dd )

library( arm )
display(M1)
display(M2)
 
library( texreg )
screenreg( list( M1, M2, M3 ) )
```
