---
title: "Pretty ANOVA Tables with `kable`"
author: "Luke Miratrix and Josh Gilbert"
editor: 
  markdown: 
    wrap: sentence
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

## R Setup

We load the `tidyverse` and `knitr`.
The `kable` function from `knitr` makes our tables look nice!

```{r}
library(tidyverse)
library(knitr)
library(broom)
```

## Create fake data

We create a data set called `a` that has 100 observations and specifies our outcome `Y` as a funciton of two uncorrelated variables `A` and `B`

```{r}
a <- tibble( A = rnorm( 100 ),
            B = rnorm( 100 ),
            Y = A * 0.2 + B * 0.5 + rnorm( 100, 0, 1 ) )
```

## Run the Models

We fit two models, one with `A` and `B`, the other with just `A`.

```{r}
M1 <- lm( Y~ A + B, data = a )
M2 <- lm( Y ~ A, data = a )
```

## Comparing the Models

We use the `anova` function to compare the two models (see also the chapter on Likelihood Ratio tests).
We see that `B` improves the model fit significantly.

```{r}
aa = anova( M2, M1 )
aa

aa |> 
  tidy() |> 
  kable()
```

## Compare to the Significance test on `B`

Note that the p value for `B` is identical to the ANOVA results above.
Why bother with ANOVA?
It can test more complex hypotheses as well (multiple coefficients, random effects, etc.)

```{r}
M1 |> 
  tidy() |> 
  kable()
```
