---
title: "Connecting the three dots: An HSB Model"
author: "Luke Miratrix"
editor: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# various libraries
library( arm )
library( foreign )
library( tidyverse )
library( lme4 )

# read student data
dat = read.spss( "data/hsb1.sav", to.data.frame=TRUE )
head( dat )

# read school data
sdat = read.spss( "data/hsb2.sav", to.data.frame=TRUE )
head( sdat )


dat = merge( dat, sdat, by="id", all.x=TRUE )

head( dat )


# Fitting a baseline model with ses 
M0 = lmer( mathach ~ 1 + ses + (1|id), data=dat )
display( M0 )


#  ** STOP HERE FOR LECTURE ** 



```

This handout shows (1) a mathematical model, (2) the `lmer` syntax for that model, and (3) the output for that model, for the model discussed in Lecture 2.4.
This handout is designed to help translate between these three different worlds.

## The mathematical model

Level 1 models: $$
\begin{aligned}
y_{ij} &= \beta_{0j} + \beta_{1j} ses_{ij} + \beta_2 female_{ij} +  \epsilon_{ij} \\
\epsilon_{ij} &\sim N( 0, \sigma^2_y ) \\
\end{aligned}
$$

Level 2 models: $$
\begin{aligned}
\beta_{0j} &= \gamma_{00} + \gamma_{01} sector_j + \gamma_{02} meanSES_j + u_{0j} \\
\beta_{1j} &= \gamma_{10} + \gamma_{11} sector_j + u_{1j} 
\end{aligned}
$$ with $$
\begin{pmatrix} u_{0j} \\
u_{1j}
\end{pmatrix} \sim  N
\begin{bmatrix}
\begin{pmatrix}
0 \\
0
\end{pmatrix}\!\!,&
\begin{pmatrix}
\tau_{00} & \tau_{01}\\
 & \tau_{11} \\
\end{pmatrix}
\end{bmatrix} .
$$

The $\tau_{01}$ is the covariance of the random intercept and random slope.
We usually look at the correlation of

$$ \rho = \frac{ \tau_{01} }{ \sqrt{ \tau_{00} \tau_{11} } } . $$

The estimated $\rho$ is what R gives us in the printed output, rather than $\tau_{01}$.

The derivation of the reduced form is:

$$
\begin{aligned}
y_{ij} &= \beta_{0j} + \beta_{1j} ses_{ij} + \epsilon_{ij}\\
&= \left( \gamma_{00} + \gamma_{01} sector_j + \gamma_{02} meanSES_j + u_{0j} \right)+ (\gamma_{10} + \gamma_{11} sector_j + u_{1j}) ses_{ij} + \beta_2 female_{ij} + \epsilon_{ij} \\
&= \gamma_{00} + \gamma_{01} sector_j + \gamma_{02} meanSES_j  + u_{0j}  + \gamma_{10}ses_{ij} + \gamma_{11} sector_j ses_{ij} + u_{1j} ses_{ij} + \beta_2 female_{ij} +  \epsilon_{ij}  \\
&= \gamma_{00} + \gamma_{01} sector_j + \gamma_{02} meanSES_j +  \gamma_{10}ses_{ij} + \gamma_{11} sector_j ses_{ij} + \beta_2 female_{ij}  + \left(u_{0j} + u_{1j} ses_{ij} + \epsilon_{ij} \right) 
\end{aligned}
$$ This formula is what we will give to `lmer()` in R's formula notation.

## The lmer code

```{r}
M1 = lmer( mathach ~ 1 + female + ses*sector + 
             meanses + (1+ses|id),
           data = dat )
```

This code is the exact same model, using the fact that `ses*sector` means `ses + sector + ses:sector`.
I.e., the above is exactly the same as this more explicitly written R code:

```{r}
M1 = lmer( mathach ~ 1 + sector + meanses + ses + sector:ses + female + (1+ses|id),
           data = dat )
```

Each term in the expanded formula corresponds to a math symbol in the mathematical model.
The `(1+ses|id)` make our random effects, and tie to all the $\tau$ terms.
The residual variance $\sigma^2_y$ is the only parameter not explicitly listed in the above model.

## The output

```{r}
display( M1 )
```

Now, using this output, we have estimates for all our mathematical modeling parameters:

-   $\gamma_{00} = 12.79$ - The overall average math achievement for a student with 0 ses in a public school with 0 mean SES.
-   $\gamma_{01} = 1.29$ - The average difference between otherwise equivilent catholic and public schools.
-   $\gamma_{02} = 3.04$ - The impact on average achievement due to mean SES of schools. Higher SES schools have higher achievement.
-   $\gamma_{10} = 2.73$ - The average slope of ses vs. math achievement in public schools.
-   $\beta_2 = -1.18$ - The gender gap; girls have lower math scores on average.
-   $\gamma_{11} = -1.31$ - The difference in slope between public and catholic schools (catholic schools have flatter slopes).
-   $\tau_{00} = 1.45^2$ - Variation in overall intercept of schools (within category of public or catholic, and beyond mean SES).
-   $\tau_{11} = 0.18^2$ - The variation in the random slopes for ses vs. math achievement.
-   $\rho = 0.65$ - The random intercepts are correlated with random slopes. High achievement schools have more discrepancy between low and high ses students.
-   $\sigma_y = 6.05$ - The unexplained student variation within school.
