---
title: "Using sjPlot to plot models"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(lme4)
library(sjPlot)
library(ggeffects)
library(tibble)
library(dplyr)
library( tidyverse )
```

An alternative tool for making plots of various models is to use the `sjPlot` package. Below are some examples of how to use `sjPlot` to visualize mixed-effects models fitted with `lme4`.

We first fit a few models to some fake data:

```{r simulate-models, include=FALSE}
set.seed(123)
df <- tibble(
  cluster = rep(1:50, each = 10),
  treatment = rbinom(500, 1, 0.5),
  cov1 = rnorm(500),
  female = rbinom(500, 1, 0.5),
  SPED = rbinom(500, 1, 0.2),
  Essay = rnorm(500, 70, 10),
  Y = rnorm(500)
) |> 
  mutate(
    Y = 1 + 0.5 * treatment + 0.3 * cov1 + 0.4 * treatment * cov1 + rnorm(500),
    cluster = factor(cluster),
    female = factor(female)
  )
```

```{r}
head( df )

m1 <- lmer(Y ~ treatment + (1 | cluster), data = df)
m2 <- lmer(Y ~ treatment + cov1 + (1 | cluster), data = df)
m3 <- lmer(Y ~ treatment * cov1 + (1 | cluster), data = df)
```

We next look at some of the characteristics of these models and data, using `sjPlot`. We load the library:

```{r}
library(sjPlot)
```

Now off we go!

## Frequency Plot for Binary Variable

```{r plot-sped}
plot_frq(df$SPED)
```

## Boxplot by Group

`sjPlot` has some general data viz stuff:

```{r plot-essay-by-female}
plot_grpfrq(var.cnt = df$Essay, var.grp = df$female, type = "box")
```

## Fixed Effects Plot

Coefficent plots (see prior chapter):

```{r fixed-effects}
plot_model(m3, type = "est")
```

## Marginal Effects (Main Effects)

```{r marginal-effects}
plot_model(m2, type = "eff", terms = c("cov1", "treatment"))
```

## Interaction Effects Plot

```{r interaction-effects}
plot_model(m3, type = "int", terms = c("cov1", "treatment"))
```

## Random Slopes Plot

```{r random-slopes}
plot_model(m3, type = "re", sort.est = TRUE) 
```

## Predicted Values by Covariate

```{r predicted-cov1}
plot_model(m3, type = "pred", terms = "cov1")
```

## ggeffects Plot of Marginal Effects

```{r ggeffects-plot}
plot(ggpredict(m3, terms = c("cov1", "treatment")))
```
