---
title: "Within, Between, and Contextual Effects"
author: "Josh Gilbert"
editor: 
  markdown: 
    wrap: sentence
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

Many find it hard to keep track of within, between, and contextual effects in MLMs.
This short walkthrough shows how to fit and interpret each model using the HSB data.

```{r}
# load libraries
library(tidyverse)
library(lme4)
library(sjPlot)
library(ggeffects)
library(haven)

# clear memory
rm(list = ls())

select <- dplyr::select

# load HSB data
hsb <- read_dta("data/hsb.dta") |> 
  select(mathach, ses, schoolid) |> 
  group_by(schoolid) |> 
  mutate(grp_mean_ses = mean(ses)) |> 
  ungroup() |> 
  mutate(grp_center_ses = ses - grp_mean_ses)
```

## Fitting the Models

```{r}
ols <- lm(mathach ~ ses, hsb)
fe <- lm(mathach ~ ses + factor(schoolid), hsb)
ri <- lmer(mathach ~ ses + (1|schoolid), hsb)
ri_within <- lmer(mathach ~ grp_center_ses + (1|schoolid), hsb)
ri_between <- lmer(mathach ~ grp_mean_ses + (1|schoolid), hsb)
re_wb <- lmer(mathach ~ grp_center_ses + grp_mean_ses + (1|schoolid), hsb)
contextual <- lmer(mathach ~ ses + grp_mean_ses + (1|schoolid), hsb)

tab_model(ols, fe, ri, ri_within, ri_between, re_wb, contextual,
          p.style = "stars",
          show.ci = FALSE,
          show.se = TRUE,
          keep = "ses",
          show.dev = TRUE,
          dv.labels = c("OLS",
                        "Fixed Effects",
                        "Rand. Int.",
                        "RI Within",
                        "RI Between",
                        "REWB",
                        "Mundlak"))
```

## Interpretation

### OLS

Ignoring school membership, students who are 1-unit higher in SES are predicted to score 3.18 points higher in math

### Fixed Effects

Holding constant school, students who are 1-unit higher in SES are predicted to score 2.19 points higher in math

### Random Intercepts

Students who are 1-unit higher in SES are predicted to score 2.39 points higher in math; schools that are 1-unit higher in mean SES are predicted to have mean math scores 2.39 points higher.

(A precision-weighted average of the within and between effects; Within: 2.19, Between: 5.86. If the RE assumption holds, these are the same in the population so we get more precision by averaging them together. However, in social science, they are rarely the same!)

### Random Intercepts, Within Effect

Holding constant school, students who are 1-unit higher in SES are predicted to score 2.19 points higher in math.
This is the same coefficient as the FE model, but in an RI framework.
We have "controlled for school" manually by demeaning the SES variable.

### Random Intercepts, Between

Schools that are 1-unit higher in mean SES are predicted to have mean math scores 5.86 points higher.

### Random Effects within and Between

Holding constant school, students who are 1-unit higher in SES are predicted to score 2.19 points higher in math; schools that are 1-unit higher in mean SES are predicted to have mean math scores 5.86 points higher.
We get the within and between effects in a single model!

### Contextual/Mundlak

Holding constant school, students who are 1-unit higher in SES are predicted to score 2.19 points higher in math; *holding constant student SES*, a student that attends a school with 1-unit higher in mean SES are predicted to have mean math scores 3.68 points higher.
The contextual effect is the *difference* in the within and between effects, and its significance test allows us to determine if it is necessary.
Mathematically, the Mundlak model and REWB are *identical*, as you can see from the deviance statistics.
You would choose one over the other depending on your preferred interpretation.

## Further Reading

[@antonakis2019a]
